\documentclass[11pt]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\signature{Rylan Malarchick\\
Department of Engineering Physics\\
Embry-Riddle Aeronautical University\\
\texttt{malarchr@erau.edu}}

\address{Rylan Malarchick\\
Department of Engineering Physics\\
Embry-Riddle Aeronautical University\\
1 Aerospace Boulevard\\
Daytona Beach, FL 32114}

\begin{document}

\begin{letter}{Editor-in-Chief\\
IEEE Transactions on Quantum Engineering}

\opening{Dear Editor,}

I am pleased to submit our manuscript entitled \textbf{``Verified Implementation of GRAPE Pulse Optimization for Quantum Gates with Hardware-Representative Noise Models''} for consideration in IEEE Transactions on Quantum Engineering.

\textbf{Summary:} This work presents QubitPulseOpt, an open-source Python framework for quantum optimal control that bridges the critical ``sim-to-real'' gap in pulse optimization. The framework implements GRAPE (Gradient Ascent Pulse Engineering) with Lindblad master equation simulation using hardware-representative parameters from IQM's Garnet quantum processor. Key contributions include:

\begin{itemize}
    \item Demonstration of 99.14\% gate fidelity in closed-system simulation, representing 77$\times$ error reduction versus Gaussian baselines
    \item An 864-test verification suite (74\% code coverage) adhering to NASA JPL Power-of-10 safety-critical coding standards
    \item API connectivity infrastructure for hardware-calibrated optimization workflows
    \item Open-source release enabling reproducible quantum control research
\end{itemize}

\textbf{Relevance to IEEE TQE:} This manuscript addresses the intersection of quantum control theory and software engineering---a topic central to TQE's scope. As quantum hardware matures, the need for rigorously verified control software becomes critical. Our framework provides both theoretical contributions (hardware-aware pulse optimization) and practical tools (production-grade open-source code).

\textbf{Prior Publication:} An earlier version of this work appeared as arXiv preprint 2511.12799 (November 2025). This submission represents a substantially expanded and refined manuscript suitable for peer review.

\textbf{Suggested Reviewers:}
\begin{itemize}
    \item Prof. Steffen Glaser (Technical University of Munich) -- quantum optimal control expert
    \item Prof. Christiane Koch (Freie Universität Berlin) -- open quantum systems
    \item Dr. Sarah Sheldon (IBM Quantum) -- superconducting qubit pulse engineering
\end{itemize}

I confirm that this manuscript has not been published elsewhere and is not under consideration by another journal. All authors have approved the manuscript and agree with submission to IEEE TQE.

Thank you for considering our submission.

\closing{Sincerely,}

\end{letter}
\end{document}
